Accelerated Inference for Partially Observed Stochastic Processes using Automatic Differentiation

### 2024-06-13 paper/ms.tex
update from overleaf

Scibior and Wood \cite{scibior21} showed that the estimators derived by Poyiadjis et al. \cite{poyiadjis11} can be attained with standard AD software, using an algorithmic procedure called a ``stop-gradient'' which allows selected exressions to be evaluated but not differentiated.
Unfortunately, \cite{scibior21} used their estimator within an algorithm which does not have the plug-and-play property and which has high Monte Carlo variability.
We use \cite{scibior21} and \cite{poyiadjis11} as our starting point, while developing a new approach which remedies the weaknesses of these papers.

@@

MOP-$\alpha$ avoids the discontinuity that is problematic for the standard particle filter.
This makes MOP-$\alpha$ convenient for the application of AD.
Yet, MOP-$\alpha$  can nevertheless provide a consistent log-likelihood estimate.

We derive a linear convergence rate for stochastic gradient descent (SGD) using MOP-$\alpha$ in the  presence of strong convexity.
Critically, the estimator is effective within a neighborhood of the maximum but is relatively slow at reaching this neighborhood.
That behavior is complementary to that of the iterated filtering algorithm \cite{ionides06-pnas,ionides15} which provides a relatively fast and stable way to identify this neighborhood.
We therefore build a hybrid algorithm called {\it Iterated Filtering with Automatic Differentiation} (IFAD) that warm-starts first-order or second-order gradient methods with a preliminary solution obtained from a few rounds of iterated filtering.
Promising numerical results indicate that IFAD beats IF2 (and by the numerical results of \cite{ionides15}, also IF1 and the Liu-West filter \cite{liuwest01}) on a challenging problem in epidemiology, the Dhaka cholera model of \cite{king08}.

These improvements also extend to Bayesian inference, as we show in the supplement that we can use the MOP-$\alpha$ gradient estimates within a No-U-Turn Sampler (NUTS) \cite{homan14} in conjunction with a nonparametric empirical Bayes-style prior estimated with IF2 to reduce the burn-in period of particle MCMC \cite{andrieu10} from the $1.4 \times 10^6$ iterations in \cite{fasiolo16} to just $500$.

@@

Directly differentiating through a particle filter, for a fixed value of $\omega$, leads to a biased gradient estimator due to the discreteness of the resampling operation \cite{corenflos21}.
We avoid this issue by constructing a particle filtering algorithm called {\it Measurement Off-Parameter} with discount factor $\alpha$ (MOP-$\alpha$), defined by the pseudocode in Algorithm~\ref{alg:mop}.
For MOP-$\alpha$, the resampling indices are invariant to $\theta$ when $\omega$ and $\phi$ are fixed.
We assume that the measurement model is differentiable in $\theta$.
Further, we assume that the simulator is a differentiable function of $\theta$ for every fixed $\omega$, a condition that requires the latent process to be a continuous random variable.
\ed{WE COULD CALL THIS THE REPARAMETERIZATION TRICK, BUT HERE FOR AD IT ARISES THROUGH THE STRUCTURE OF THE CODE SO WE DO NOT HAVE TO EXPLICITLY THINK OF THE EXISTENCE OF A REPARAMETERIZATION.}
Under these assumptions, direct differentiation of MOP-$\alpha$ is available via AD.
As we explain subsequently, MOP-$\alpha$ is constructed so this direct derivative obtains the score estimator of  \cite{poyiadjis11, scibior21} when $\alpha=1$ and that of \cite{naesseth18} when $\alpha=0$.
Setting $\alpha<1$ adds bias but reduces variance, raising opportunities for a favorable tradeoff.

@@

\subsection{Algorithm Outline} 
We obtain two coupled sets of particles, one under $\phi \in \Theta$, and another with the process model at $\theta \in \Theta$ but with the resampling indices constructed from the first pass, with a ``baseline parameter'' $\phi$.
As the process model follows $\theta$ but the measurement model follows the baseline parameter $\phi$, we therefore call this a ``measurement off-parameter'' particle filter.
The resampling indices, which we write as $k_j \sim \text{Categorical}(g^{\phi}_{n,1},...,g^{\phi}_{n,J})$, are a function of $\phi$ for any value of $\theta$.
For the categorical distribution, we use systematic resamplig \cite{arulampalam02,king16} which usually has superior performance to multinomial resampling.
If $\theta$ and $\phi$ coincide, one only needs one particle filter run at $\theta=\phi$, otherwise one needs two runs at the same seed $\omega \in \Omega$.

@@

This yields a suitably reweighted particle filter that targets the posterior under $\theta$ and produces a strongly consistent likelihood estimate for the likelihood at $\theta$. Taking the gradient with respect to $\theta$ of the estimate for the log-likelihood at $\theta$ then yields an estimate for the score at $\theta$.

@@

The correction factors, $w^{P,\theta}_{n,j}$ and $w^{F,\theta}_{n,j}$, accumulate as $n$ increases, leading to numerical instability.
One can discount the correction factors from previous timesteps by some $\alpha \in [0,1]$ as in Equation \ref{eq:weighting-scheme} while still targeting the posterior and likelihood if $\theta=\phi$.
This lets us optimize a bias-variance tradeoff for the MOP-$\alpha$ score estimate.
When $\alpha=1$, MOP-$1$ maintains the memory of each particle's ancestral trajectory, and Theorem~\ref{thm:mop-functional-forms} shows that it recovers the consistent but high-variance gradient estimate from \cite{poyiadjis11, scibior21}.
When $\alpha=0$, MOP-$0$ considers only single-step transition dynamics, recovering the low-variance but asymptotically biased gradient estimator of \cite{naesseth18}.

@@

\cite{scibior21} show that the estimate of \cite{naesseth18} is the gradient of a vanilla particle filter when resampling terms are dropped, and also recover the estimate of \cite{poyiadjis11} through applying a ``stop-gradient trick" to the particle filter. It turns out that both of these, when applied on the bootstrap filter, correspond to special cases of MOP-$\alpha$.

@@

If $\theta$ is evaluated at $\phi$, the particles at $\theta$ and $\phi$ coincide. One then only needs to run one particle filter at $\theta=\phi$, setting the particles at $\phi$ to be copies of the particles at $\theta$ where gradients don't propagate. This is done algorithmically through the \texttt{stop\_gradient()} function in \texttt{JAX}, providing a mathematical justification for the ``stop-gradient trick" of \cite{scibior21}.

@@

The  convergence of IF2 (and so the first stage of IFAD) to a neighborhood of the MLE happens fairly quickly in practice. In the case of the Dhaka cholera model of \cite{king08}, when an aggressive geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, initial convergence happens within 40 iterations. Finding the MLE itself with IF2, however, takes much longer, as one has to use a less aggressive cooling rate to do so. For example, \cite{ionides15} use 100 iterations with the Dhaka cholera model of \cite{king08}, while \cite{wheeler23} use 200 with Model 1 in \cite{lee20}. By simply requiring that the first stage of IFAD get to a neighborhood of the MLE and not the MLE itself, we are able to substantially reduce the number of IF2 iterations required.

@@

On the other hand, the second stage of IFAD enjoys a linear convergence rate under the usual strongly-convex and smoothness assumptions in convex optimization, as we show below in Theorem \ref{thm:mop-convergence}.

@@

\begin{figure}
    \centering
    \includegraphics[width=\arxiv{12cm}{\textwidth/2}]{../imgs/095/tikzcholera.png}
    \vspace*{-7mm}
    \caption{Illustration of Dhaka cholera model from \cite{king08}. \ed{CHANGE H TO P? ALSO, THERE ARE ADVANTAGES TO PUTTING THE TIKZ SOURCE CODE INTO THE MS, TO ALLOW FOR FINE-TUNING EDITS.}}
    \label{fig:tikz-cholera}
\end{figure}

@@

We tested IFAD against IF2 on a global search problem for the Dhaka cholera model.
We re-implemented IF2 in order to do so, but we compare our results with the results of \cite{ionides15} (labeled "IF2 2015").
Our re-implementation ourperforms that of \cite{ionides15}, likely due to a better choice of algorithmic parameters.
For each method, we performed 100 searches, initialized with 100 initial starting parameter vectors drawn uniformly from the same wide bounding box used in \cite{ionides15}. We summarize our findings below.

@@

\ed{PERHAPS THE 2ND D.P. OF SIGNIFICANCE CAN BE AVOIDED? ALSO, IF WE'RE GOING TO ADVISE MOP-ALPHA FOR $0<\alpha<1$, THAT COULD BE SHOWN HERE.?}

\begin{table}[h!]
\centering
\input{../imgs/095/table}
\caption{Maximum log-likelihood found by IF2, IFAD, and MOP alone. IFAD performs the best among all methods. Our implementation of IF2 outperforms that of \cite{ionides15}, but still ultimately underperforms IFAD. IFAD manages to find the MLE, matching the highest log-likelihood previously found in the \texttt{dacca()} model implemented within the \texttt{pomp} package of \cite{king16}.}
\label{table:mle}
\end{table}

@@

\begin{figure}[htbp!]
    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{../imgs/095/pairs.png}
    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{../imgs/095/qq.png}
    \caption{Scatterplots depicting the performance of IFAD against that of IF2. \textbf{Left:} Paired searches from the same starting point. Controlling for initial starting point, tuning $\alpha$ allows IFAD-$0.97$ to strictly improve on IF2, \cite{poyiadjis11}, and \cite{naesseth18}, on almost every iteration. \textbf{Right:} Q-Q plot of ranked IFAD searches against ranked IF2 searches. It is clear that on average, IFAD performs best, and manages to find the MLE while no IF2 search successfully gets within 7 log-likelihood units of it. We use a dotted red line to display the MLE in both. }
    \label{fig:scatter}
\end{figure}

@@

\begin{figure}[ht]
    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{../imgs/095/boxplot.png}
    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{../imgs/095/boxplot_all.png}
    \caption{\textbf{Left:} Raincloud plot depicting the performance of IFAD and IF2 where we plot the results of the best run out of every ten runs, simulating the result of the common procedure of running a few searches and choosing the best one. \textbf{Right:} Raincloud plot of all searches. IFAD outperforms all other methods, and the gradient steps improve on the warm-start given by running 40 IF2 iterations.
    We use a dotted red line to display the MLE.}
    \label{fig:boxplot}
\end{figure}

@@

\ed{IS THE IFAD(0.95) MEDIAN LINE IN FIG 4 (RIGHT) HIDING UNDER IFAD(1)? IF NOT, WEHRE IS IT?}

@@

Here, we will show that MOP-$\alpha$ targets the posterior and likelihood and yields the estimators of \cite{poyiadjis11, scibior21, naesseth18} as special cases when differentiated, and characterize rates for its bias and variance under different choices of $\alpha$. To do so, we require the following assumptions:

@@

\subsection{MOP-$\alpha$ Targets the Posterior}

We show here that MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood when $\alpha=1$ or $\theta=\phi$.
Employing $\alpha<1$ lead to inconsistency when $\theta$ deviates from $\phi$, but this bias disappears as $\theta$ approaches $\phi$.

@@

\begin{thm}[MOP-$\alpha$ Targets the Posterior and Likelihood]
    \label{thm:mop-targeting}
    When $\alpha=1$ or $\theta=\phi$, MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood. That is, for $\pi_n(\theta)=f_{X_{1:n}|Y_{1:n} ; \theta}$ and any measurable and bounded functional $h$ and for $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$ or $\prod_{n=1}^N L_n^{B,\theta,\alpha}$, it holds that
    \vspace*{-2.5mm}
    \begin{equation}
        \frac{\sum_{j=1}^J h(x_{n,j}^{A,F, \theta}) w_{n,j}^{A,F,\theta}}{\sum_{j=1}^J w_{n,j}^{A,F,\theta}} \stackrel{a.s.}{\to} E_{\pi_n(\theta)} h, \;\; \hat\lik(\theta)  \stackrel{a.s.}{\to} \lik(\theta).
    \vspace*{-2.5mm}
    \end{equation}
\end{thm}

@@

\begin{proof}
    We provide a proof sketch here, deferring most of the details, and discussion of the after-resampling conditional likelihood estimate $L_n^{A,\theta,\alpha}$, to the supplementary information. 
    When $\theta=\phi$, regardless of the value of $\alpha$, the ratio ${g_{n,j}^\theta}/{g_{n,j}^\phi}=1,$ and this reduces to the vanilla particle filter.
    When $\alpha=1$ and $\theta\neq\phi,$ suppose inductively that $\{(X^{F,\theta}_{n-1,j},w^{F,\theta}_{n-1,j})\}_{j=1}^J$ targets $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.
    It can then be shown that $\{(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j})\}_{j=1}^J$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$, that $\{(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} g^\theta_{n,j} )\}_{j=1}^J$ targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$, and that weighting the particles by $(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}) = (X^{P,\theta}_{n,a(j)}, w^{P,\theta}_{n,a(j)} g^\theta_{n,a(j)}/ g^\phi_{n,a(j)})$ where $a(\cdot)$ is the ancestor function,
    resampling with probabilities proportional to $g^\phi_{n,j}$, also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
    If the likelihood is estimated with the before-resampling conditional likelihoods $\hat\lik(\theta) = \prod_{n=1}^N L_n^{B,\theta,\alpha}$, the strong consistency is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$. 
\end{proof}

@@

\begin{thm}[Consistency of MOP-$1$ Gradient Estimate]
    The gradient estimate of MOP-$\alpha$ when $\alpha=1$, $\theta=\phi$ is strongly consistent for the score: $\nabla_\theta \hat\ell_J^1(\theta) \stackrel{a.s.}{\to} \nabla_\theta \ell(\theta)$ as $J \to \infty$.
    \label{thm:mop-grad-consistency}
\end{thm}
\begin{proof}
    Fix $\omega \in \Omega$, and set $\phi = \theta$, where $\theta$ is the point at which we wish to evaluate the gradient. The sequence $(\nabla_\theta \hat\lik_J^1(\theta)(\omega))_{J \in \mathbb{N}}$ is uniformly bounded over all $J$ by Assumption \ref{assump:local-bounded-derivative}. Again by Assumption \ref{assump:local-bounded-derivative}, the second derivative of $\hat\lik_J^1(\theta)(\omega)|_{\theta=\theta'}$ is also bounded by $H^*$ for almost every $\omega\in \Omega$ and every $\theta'\in \Theta$. So $(\nabla_\theta \hat\lik_J^1(\theta)( \omega))_{J \in \mathbb{N}}$ is uniformly Lipschitz, and therefore uniformly equicontinuous for almost every $\omega \in \Omega$.

    By Arzela-Ascoli, there is a uniformly convergent subsequence. But there is only one subsequential limit, as we can treat the gradient estimate at $\theta$ as a bounded functional of the particles by Assumption \ref{assump:local-bounded-derivative}, allowing us to apply Theorem \ref{thm:mop-targeting} to see that the sequence $(\nabla_\theta \hat\lik_J^1(\theta, \omega))_{J \in \mathbb{N}}$ converges pointwise for $\theta=\phi$ and almost every $\omega \in \Omega$. So the whole sequence must converge uniformly to $\lim_{J \to \infty} \nabla_\theta \hat\lik_J^1(\theta)(\omega).$

>>>>

Scibior and Wood \cite{scibior21} showed that the estimators derived by Poyiadjis et al. \cite{poyiadjis11} can be attained with standard AD software, using an algorithmic procedure called a ``stop-gradient'' which allows selected expressions to be evaluated but not differentiated.
Unfortunately, \cite{scibior21} used their estimator within an algorithm which does not have the plug-and-play property and which has high Monte Carlo variability, while the ``stop-gradient'' procedure is introduced without mathematical motivation.
We use \cite{scibior21} and \cite{poyiadjis11} as our starting point, while developing a new approach which remedies the weaknesses of these papers.

@@

MOP-$\alpha$ avoids the discontinuity that is problematic for the standard particle filter in a novel way.
This particle filter is smooth by construction, allowing us to bypass the seemingly incompatible paradox of differentiating through a Monte Carlo algorithm -- by instead using AD to differentiate through a series of measurement density ratios and a differentiable simulator.
Yet, MOP-$\alpha$ can nevertheless provide a consistent log-likelihood estimate.

We derive a linear convergence rate for stochastic gradient descent (SGD) using MOP-$\alpha$ in the  presence of strong convexity.
Critically, the estimator is effective within a neighborhood of the maximum but is relatively slow at reaching this neighborhood.
That behavior is complementary to that of the iterated filtering algorithm \cite{ionides06-pnas,ionides15} which provides a relatively fast and stable way to identify this neighborhood.
We therefore build a simple but effective hybrid algorithm called {\it Iterated Filtering with Automatic Differentiation} (IFAD) that warm-starts first-order or second-order gradient methods with a preliminary solution obtained from a few rounds of iterated filtering.
Promising numerical results indicate that IFAD beats IF2 (and by the numerical results of \cite{ionides15}, also IF1 and the Liu-West filter \cite{liuwest01}) on a challenging problem in epidemiology, the Dhaka cholera model of \cite{king08}.

These improvements also extend to Bayesian inference, as we show in Section \ref{sec:bayes} that we can use the MOP-$\alpha$ gradient estimates within a No-U-Turn Sampler (NUTS) \cite{homan14} in conjunction with a nonparametric empirical Bayes-style prior estimated with IF2 to reduce the burn-in period of particle MCMC \cite{andrieu10} on the Dhaka cholera model of \cite{king08} from the $700,000$ iterations in \cite{fasiolo16} to just $500$. To the best of our knowledge, attaining rapid mixing on this challenging problem has not previously been achievable.

@@

Directly differentiating through a particle filter, for a fixed value of $\omega$, leads to a biased gradient estimator due to the discreteness of the resampling operation \cite{corenflos21}.
We avoid this issue by constructing a particle filtering algorithm called {\it Measurement Off-Parameter} with discount factor $\alpha$ (MOP-$\alpha$), defined by the pseudocode in Algorithm~\ref{alg:mop}. In short, instead of differentiating through a Monte Carlo algorithm directly, we instead use AD to differentiate through a series of measurement density ratios and a differentiable simulator.

This algorithm is an example of a new class of algorithms we call ``off-parameter'' particle filters that resample the particles according to an arbitrary resampling rule that is not necessarily in proportion to the target distribution of interest, while maintaining weights that encode the cumulative discrepancy between the resampling distribution and the target distribution. 
We later show in Theorem \ref{thm:mop-targeting} that off-parameter particle filters successfully target the filtering distribution when the correction is sufficient.

In the case of MOP-$\alpha$, we wish to evaluate the likelihood at some $\theta \in \Theta$, but instead resample the particles according to the indices generated by a vanilla particle filter run at some $\phi \in \Theta$. This ensures that the resampling indices are invariant to $\theta$ when $\omega$ and $\phi$ are fixed, bypassing the issue of Monte Carlo resampling. 
When the measurement model is differentiable in $\theta$ and the simulator is a differentiable function of $\theta$ for every fixed $\omega$ (a condition that requires the latent process to be a continuous random variable, also known as the reparameterization trick \cite{corenflos21}), direct differentiation of MOP-$\alpha$ is available via AD.
As we explain subsequently, MOP-$\alpha$ is constructed so this direct derivative obtains the score estimator of  \cite{poyiadjis11, scibior21} when $\alpha=1$ and that of \cite{naesseth18} when $\alpha=0$.
Setting $\alpha<1$ adds bias but reduces variance, raising opportunities for a favorable tradeoff as illustrated in theory (Theorem \ref{thm:mop-biasvar}) and in practice (Figure \ref{fig:biasvar}).

@@

\subsection{Algorithm Outline} 
We obtain two coupled sets of particles, one under $\phi \in \Theta$, and another with the process model at $\theta \in \Theta$ but with the resampling indices constructed from the first pass, with a ``baseline parameter'' $\phi$.
As the process model follows $\theta$ but the measurement model follows the baseline parameter $\phi$, we therefore call this a ``measurement off-parameter'' particle filter.
The resampling indices, which we write as $k_j \sim \text{Categorical}(g^{\phi}_{n,1},...,g^{\phi}_{n,J})$, are a function of $\phi$ for any value of $\theta$.
For the categorical distribution, we use systematic resampling \cite{arulampalam02,king16} which usually has superior performance to multinomial resampling.
If $\theta$ and $\phi$ coincide, one only needs one particle filter run at $\theta=\phi$, otherwise one needs two runs at the same seed $\omega \in \Omega$.

@@

This yields a suitably reweighted particle filter that targets the filtering distribution under $\theta$ and produces a strongly consistent likelihood estimate for the likelihood at $\theta$. As the resampling indices are invariant to $\theta$ when $\omega$ and $\phi$ are fixed, we successfully bypass the issue of Monte Carlo resampling when the simulator and and measurement model are differentiable. Taking the gradient with respect to $\theta$ of the log-likelihood estimate for $\theta$ then yields an estimate for the score at $\theta$.

@@

The correction factors, $w^{P,\theta}_{n,j}$ and $w^{F,\theta}_{n,j}$, accumulate as $n$ increases, leading to numerical instability.
One can discount the correction factors from previous timesteps by some $\alpha \in [0,1]$ as in Equation \ref{eq:weighting-scheme} while still targeting the filtering distribution and likelihood if $\theta=\phi$.
This lets us optimize a bias-variance tradeoff for the MOP-$\alpha$ score estimate.
When $\alpha=1$, MOP-$1$ maintains the memory of each particle's ancestral trajectory, and Theorem~\ref{thm:mop-functional-forms} shows that it recovers the consistent but high-variance gradient estimate from \cite{poyiadjis11, scibior21}.
When $\alpha=0$, MOP-$0$ considers only single-step transition dynamics, recovering the low-variance but asymptotically biased gradient estimator of \cite{naesseth18}. 
This discount factor is novel and effective both in theory (Theorem \ref{thm:mop-biasvar}) and in practice (Figure \ref{fig:biasvar}).

@@

\cite{scibior21} show that the estimate of \cite{naesseth18} is the gradient of a vanilla particle filter when resampling terms are dropped, and also recover the estimate of \cite{poyiadjis11} through the use of a stop-gradient operation in the AD procedure. It turns out that both of these, when applied on the bootstrap filter, correspond to special cases of MOP-$\alpha$.

@@

If $\theta$ is evaluated at $\phi$, the particles at $\theta$ and $\phi$ coincide. One then only needs to run one particle filter at $\theta=\phi$, setting the particles at $\phi$ to be copies of the particles at $\theta$ where gradients don't propagate. This is done algorithmically through the \texttt{stop\_gradient()} function in \texttt{JAX}, providing a mathematical justification for the use of the stop-gradient operation by \cite{scibior21}.

@@

The  convergence of IF2 (and so the first stage of IFAD) to a neighborhood of the MLE happens rapidly in practice. 
In the case of the Dhaka cholera model of \cite{king08}, when an aggressive geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, initial convergence happens within 40 iterations. 
Finding the MLE itself with IF2, however, takes much longer, as one has to use a less aggressive cooling rate to do so. 
For example, \cite{ionides15} use 100 iterations with the Dhaka cholera model of \cite{king08}, while \cite{wheeler23} use 200 with Model 1 in \cite{lee20}. 
By simply requiring that the first stage of IFAD get to a neighborhood of the MLE and not the MLE itself, we are able to substantially reduce the number of IF2 iterations required.

@@

On the other hand, the second stage of IFAD enjoys a linear convergence rate under the usual smoothness and strong convexity assumptions in the theory of convex optimization, as we show below in Theorem \ref{thm:mop-convergence}.

@@

\begin{figure}
    \centering
    \includegraphics[width=\arxiv{12cm}{\textwidth/2}]{../imgs/095/tikzcholera.png}
    \vspace*{-7mm}
    \caption{Illustration of Dhaka cholera model from \cite{king08}.}
    \label{fig:tikz-cholera}
\end{figure}

@@

We tested IFAD against IF2 on a global search problem for the Dhaka cholera model.
We re-implemented IF2 to do so, but we compare our results with the results of \cite{ionides15} (labeled "IF2 2015").
Our re-implementation ourperforms that of \cite{ionides15}, likely due to a better choice of algorithmic parameters.
For each method, we performed 100 searches, initialized with 100 initial starting parameter vectors drawn uniformly from the same wide bounding box used in \cite{ionides15}. We summarize our findings below.

@@

\begin{table}[h!]
\centering
\input{../imgs/095/table}
\caption{Maximum log-likelihood found by IF2, IFAD, and MOP alone. IFAD performs the best among all methods. Our implementation of IF2 outperforms that of \cite{ionides15}, but still ultimately underperforms IFAD. IFAD manages to find the MLE, matching the highest log-likelihood previously found in the Dhaka cholera model implemented within the \texttt{pomp} package of \cite{king16}.}
\label{table:mle}
\end{table}

@@

\begin{figure}[htbp!]
    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{../imgs/095/pairs.png}
    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{../imgs/095/qq.png}
    \caption{Scatterplots depicting the performance of IFAD against that of IF2. \textbf{Left:} Paired searches from the same starting point. Controlling for initial starting point, tuning $\alpha$ allows IFAD-$0.97$ to strictly improve on IF2, \cite{poyiadjis11}, and \cite{naesseth18}, on almost every iteration. \textbf{Right:} Q-Q plot of ranked IFAD searches against ranked IF2 searches. It is clear that on average, IFAD performs best, and manages to find the MLE while no IF2 search successfully gets within 7 log-likelihood units of it. The dotted red line shows the true maximized log-likelihood. }
    \label{fig:scatter}
\end{figure}

@@

\begin{figure}[ht]
    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{../imgs/095/boxplot.png}
    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{../imgs/095/boxplot_all.png}
    \caption{\textbf{Left:} Raincloud plot depicting the performance of IFAD and IF2 where we plot the results of the best run out of every ten runs, representing the common procedure of running a few searches and choosing the best one. \textbf{Right:} Raincloud plot of all searches. IFAD outperforms all other methods, and the gradient steps improve on the warm-start given by running 40 IF2 iterations.
    The dotted red line shows the true maximized log-likelihood.}
    \label{fig:boxplot}
\end{figure}

@@

\section{Application to Bayesian Inference}
\label{sec:bayes}

Particle MCMC, as introduced by \cite{andrieu10}, is arguably the most popular method for full-information Bayesian inference within partially observed Markov processes that does not require access to the system's transition densities. This feature enables its use in simulation-based Bayesian inference for complex scientific models where computing transition densities is not feasible, such as in disease modeling. However, the particle Metropolis-Hastings algorithm often experiences slow mixing and requires long burn-in periods. For instance, \cite{fasiolo16} reported needing $700,000$ burn-in iterations for effective sampling in a cholera model -- even after adjusting the length of each Euler timestep to be a month instead of a day to reduce computational costs.

We employ a NUTS sampler powered by a MOP-$\alpha$, with a nonparametric empirical prior initialized by the IF2 warm start from the previous section. To construct the prior, we perform a kernel density estimate (KDE) on the parameter swarm from the warm start, and use the output of that as an empirical prior (the KDE is performed to ensure some degree of mutual contiguity between the densities). This yields a prior that is roughly equivalent to a parameter cloud centered at some point in the neighborhood of the MLE, with a standard deviation of $\sqrt{(0.02\cdot0.95^{40})^2 \cdot 600}\approx 6.3\%$ of the MLE. Using a NUTS sampler powered by MOP-$\alpha$ with this prior significantly lowers the mixing time of the Markov chain to as little as $500$ iterations, as seen in Figure \ref{fig:nuts-eb}.

\begin{figure}[t!]
    \centering
    \includegraphics[width=\textwidth/\real{1.25}]{imgs/pmcmc/nuts_eb.png}
    \caption{Convergence diagnostics for a No U-Turn Sampler with the informative nonparametric empirical Bayes prior from IF2 and a KDE. We display the results for the trend parameter in the Dhaka cholera model of \cite{king08}. The NUTS sampler mixes quickly, and the posterior estimates from each chain share roughly the same posterior mode.}
    \label{fig:nuts-eb}
\end{figure}

For completeness, we note that without the empirical Bayes-style prior from IF2, NUTS with the MOP-$\alpha$ estimate alone tends to diverge. In contrast, particle Metropolis-Hastings (with or without the empirical Bayes prior from IF2) fails to effectively explore the parameter space. We display the results of the above in Figures \ref{fig:mh} and \ref{fig:nuts} within the supplementary information.

The result obtained in this section, of a speedup in the mixing time of particle MCMC by over three orders of magnitude, provides hope that particle Bayesian inference might finally be practical for the complex scientific models commonly encountered in areas like epidemiology.

@@

Here, we will show that MOP-$\alpha$ targets the filtering distribution and likelihood under $\theta$, is consistent when $\alpha=1$, and characterize rates for its bias and variance under different choices of $\alpha$. To do so, we require the following assumptions:

@@

\subsection{MOP-$\alpha$ Targets the Filtering Distribution}

We show here that MOP-$\alpha$ targets the filtering distribution under $\theta$ and is strongly consistent for the likelihood under $\theta$ when $\alpha=1$ or $\theta=\phi$.
Employing $\alpha<1$ lead to inconsistency when $\theta$ deviates from $\phi$, but this bias disappears as $\theta$ approaches $\phi$.

@@

\begin{thm}[MOP-$\alpha$ Targets the Filtering Distribution and Likelihood]
    \label{thm:mop-targeting}
    When $\alpha=1$ or $\theta=\phi$, MOP-$\alpha$ targets the filtering distribution under $\theta$ and is strongly consistent for the likelihood under $\theta$. That is, for $\pi_n(\theta)=f_{X_{n}|Y_{1:n} ; \theta}$ and any measurable and bounded functional $h$ and for $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$ or $\prod_{n=1}^N L_n^{B,\theta,\alpha}$, it holds that
    \vspace*{-2.5mm}
    \begin{equation}
        \frac{\sum_{j=1}^J h(x_{n,j}^{F, \theta}) w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{F,\theta}} \stackrel{a.s.}{\to} E_{\pi_n(\theta)} h, \;\; \hat\lik(\theta)  \stackrel{a.s.}{\to} \lik(\theta).
    \vspace*{-2.5mm}
    \end{equation}
\end{thm}

@@

\begin{proof}
    We provide a proof sketch here, deferring most of the details, and discussion of the after-resampling conditional likelihood estimate $L_n^{A,\theta,\alpha}$, to the supplementary information. 
    When $\theta=\phi$, regardless of the value of $\alpha$, the ratio ${g_{n,j}^\theta}/{g_{n,j}^\phi}=1,$ and this reduces to the vanilla particle filter.
    When $\alpha=1$ and $\theta\neq\phi,$ suppose inductively that $\{(X^{F,\theta}_{n-1,j},w^{F,\theta}_{n-1,j})\}_{j=1}^J$ targets $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.
    It can then be shown that $\{(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j})\}_{j=1}^J$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$, that $\{(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} g^\theta_{n,j} )\}_{j=1}^J$ targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$, and that weighting the particles by $(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}) = (X^{P,\theta}_{n,k_j}, w^{P,\theta}_{n,k_j} g^\theta_{n,k_j}/ g^\phi_{n,k_j})$,
    resampling the $k_j$ with probabilities proportional to $g^\phi_{n,j}$, also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
    If the likelihood is estimated with the before-resampling conditional likelihoods $\hat\lik(\theta) = \prod_{n=1}^N L_n^{B,\theta,\alpha}$, the strong consistency is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$. 
\end{proof}

@@

\begin{thm}[Consistency of MOP-$1$ Gradient Estimate]
    The gradient estimate of MOP-$\alpha$ when $\alpha=1$, $\theta=\phi$ is strongly consistent for the score: $\nabla_\theta \hat\ell_J^1(\theta) \stackrel{a.s.}{\to} \nabla_\theta \ell(\theta)$ as $J \to \infty$.
    \label{thm:mop-grad-consistency}
\end{thm}
\begin{proof}
    Fix $\omega \in \Omega$, and set $\phi = \theta$, where $\theta$ is the point at which we wish to evaluate the gradient. The sequence $(\nabla_\theta \hat\lik_J^1(\theta)(\omega))_{J \in \mathbb{N}}$ is uniformly bounded over all $J$ by Assumption \ref{assump:local-bounded-derivative}. Again by Assumption \ref{assump:local-bounded-derivative}, the second derivative of $\hat\lik_J^1(\theta)(\omega)|_{\theta=\theta'}$ is also uniformly bounded over all $J$ by $H^*$ for almost every $\omega\in \Omega$ and every $\theta'\in \Theta$. So $(\nabla_\theta \hat\lik_J^1(\theta)( \omega))_{J \in \mathbb{N}}$ is uniformly Lipschitz, and therefore uniformly equicontinuous for almost every $\omega \in \Omega$.

    By Arzela-Ascoli, there is a uniformly convergent subsequence. But there is only one subsequential limit, as we can treat the gradient estimate at $\theta$ as a bounded functional of the particles by Assumption \ref{assump:local-bounded-derivative}, allowing us to apply Theorem \ref{thm:mop-targeting} to see that the sequence $(\nabla_\theta \hat\lik_J^1(\theta)(\omega))_{J \in \mathbb{N}}$ converges pointwise for $\theta=\phi$ and almost every $\omega \in \Omega$. So the whole sequence must converge uniformly to $\lim_{J \to \infty} \nabla_\theta \hat\lik_J^1(\theta)(\omega).$
